---
title: "Exercise 7"
output: html_document
date: "2025-02-26"
---
For this exercise, you will practice detecting if there are significant differences in a continuous variable across two or more levels of catgorical (nominal and ordinal) variables. These kinds of statistical procedures are generally referred to as bivariate statistics.

1. If you need to reload the NHANES data and label the ordinal and nominal variables as factor variables, you should run the code chunk below:
```{r}
#Use this code chunk to download the NHANES 2017-2018 database and to label the nominal and ordinal variables as factor variables (if necessary):
NHANESdata <- read.csv("https://raw.githubusercontent.com/malenamalka/ANTH241-SP25/refs/heads/main/data/NHANES2017-28")
factor(c(NHANESdata$Gender, NHANESdata$Race, NHANESdata$HealthIns))
factor(c(NHANESdata$Education,NHANESdata$HHEducation,NHANESdata$HHIncome,NHANESdata$FeelBad,NHANESdata$HHFoodSec),order=TRUE)
factor(NHANESdata$BMIcat, order =TRUE, levels = c("Underweight","Healthy","Overweight","Obese"))
head(NHANESdata)
```

2. Run the code chunk below to install and load the packages that we will use in this exercise:
```{r}
library(dplyr)
library(tidyr)
library(ggplot2)
#install.packages("dunn.test")
library(dunn.test)
```

lets test whether there are age varies by gender in the NHANES sample. what is the null hypothesis? what is the alternative hypothesis?

Null hypothesis: there is no significant difference in age based on gender.
alternative hypothesis: there is an age difference that is significant based on gender

```{r}
#1.histogram of continuous variable:
hist(NHANESdatabase$Age)

#2. what are the means, standard deviations, and counts for age by gender?
NHANESdatabase%>%
  drop_na(Age)%>%
  group_by(Gender)%>%
  summarize(mean(Age), sd(Age), count=n())

#3. f-test to see if the vairance in the continous variable is the same for the different levels of the categorical vairable:
var.test(Age~Gender, data=NHANESdatabase, alternative="two.sided")

#4: run the t-test: t-test(contvar~contvar, var.equal=FALES, data=)
t.test(Age~Gender, var.equal=TRUE, data=NHANESdatabase)
```
does it refute the hypothesis: yes, there's no significant difference between age and gender. 


3. T-tests compare the variation of a continuous variable across two levels of a categorical (nominal and ordinal) variable. 

null hypothesis: no difference between bmi and gender
alt hypothesis: significant difference between BMI and gender
```{r}
#3a. What are the means, standard deviations, and counts for BMI by gender?
NHANESdatabase%>%
  drop_na(BMI)%>%
  group_by(Gender)%>%
  summarize(mean(BMI), sd(BMI), count=n())
#Let's hypothesize that there is a significant difference in BMI by gender in the NHANES 2017-2018 sample. What is the null hypothesis? What is the alternative hypothesis? We will test this hypothesis using the method shown in S&B, section 7.1.2
#1. histogram of continuous variable: hist(data$contvar)
hist(NHANESdatabase$BMI)
#2. F-test to see if the variance in the continuous variable is the same for the different levels of the categorical variable: 
#var.test(Contvar~Catvar, data=, alternative="two.sided")
var.test(BMI~Gender, data=NHANESdatabase, alternative="two.sided")

#3. According to S&B, if the variances are not equal, we should run the T-test anyways: t.test(Contvar~Catvar, var.equal=FALSE, data=)
t.test(BMI~Gender, var.equal=FALSE, data=NHANESdatabase)
```
Question: Does the above T-test support the null hypothesis or the alternative hypothesis?

yes
it does not support the alternative hypothesis because the difference is not further away from zero than 0.05.

```{r}
#3b: Does BMI differ in people with and without health insurance? What is the null hypothesis? What is the alternative hypothesis?
#the null hypothesis: BMI does not vary based on whether or not someone has health insurance. 
#the alternative hypothesis: BMI does vary based on whether or not someone has health insurance. 
#1. Find the descriptive statistics (mean, sd, and counts) of BMI by presence/absence of health insurance. 
#a. Below, I filter out levels of health insurance that are above 2 into a new database to ensure that the 7s and 9s are not part of the analysis.
NHANES_healthins <- NHANESdata%>%
  filter(HealthIns<3)
NHANES_healthins%>%
  drop_na(BMI)%>%
  group_by(HealthIns)%>%
  summarize(mean(BMI), sd(BMI), count=n())
#2. Run a test to see if the variances of the continuous variable are equal across different levels of the categorical variable:
#var.test(Contvar~Catvar, data=, alternative="two.sided")
var.test(BMI~HealthIns, data=NHANES_healthins, alternative="two.sided")
#3. We will run the T-test with the following code:
#t.test(Contvar~Catvar, var.equal=TRUE, data=)
t.test(BMI~HealthIns, var.equal=TRUE, data=NHANES_healthins)
```
Does the above T-test support or refute the alternative hypothesis?
it supports the alternative hypothesis because the difference is further away from zero than 0.05.
```{r}
#1 filter out 7777s and 9999s from sedentary (keep responses from <1321)
NHANES_sedentary <- NHANES_healthins%>%
  filter(Sedentary<1321)
#2 descriptive statistics of sedentary grouped by health insurance
NHANES_sedentary%>%
  drop_na(Sedentary)%>%
  group_by(HealthIns)%>%
  summarize(mean(Sedentary), sd(Sedentary), count=n())
#3 test to see if the variance of the continuous variable are equal across different levels of the categorical variable: var.test(contvar~catvar, data=, )
var.test(Sedentary~HealthIns, data=NHANES_sedentary, alternative="two.sided")
#4 t-test: t.test(contvar~catvar, var.equal=FALSE, data=, na.rm=TRUE)
t.test(Sedentary~HealthIns, var.equal=FALSE, data=NHANES_sedentary, na.rm=TRUE)
```

3c. Generate a hypothesis that tests if daily caloric intake differs in people who identify as female from people who identify as male. What are the null and alternative hypotheses? Run a T-test by replicating the methods above. What are your results? Do your results support or refute the alternative hypothesis?
null hypothesis: there is no significant difference based on individuals who identify as female or male compared to Caloric intake. 
alternative hypothesis: there is a significant difference based on caloric intake of individuals who identify as female or male. 
there is not a significant difference, the alternative hypothesis is refuted. 
```{r}
#2 descriptive statistics of sedentary grouped by health insurance
NHANESdatabase%>%
  drop_na(DR1TKCAL)%>%
  group_by(Gender)%>%
  summarize(mean(DR1TKCAL), sd(DR1TKCAL), count=n())
#3 test to see if the variance of the continuous variable are equal across different levels of the categorical variable: var.test(contvar~catvar, data=, )
var.test(DR1TKCAL~Gender, data=NHANESdatabase, alternative="two.sided")
#4 t-test: t.test(contvar~catvar, var.equal=FALSE, data=, na.rm=TRUE)
t.test(DR1TKCAL~Gender, var.equal=FALSE, data=NHANESdatabase, na.rm=TRUE)
```
4. Testing for significant differences in a continuous variable across three or more categorical levels. Both S&B and Kabacoff introduce you to analysis of variance (ANOVA) as the statistical method to make these comparisons. What they show you is all good. However, I would like to introduce you to a non-parametric test that is more appropriate to the NHANES data as it applies to variables that are not normally distributed. This method is similar to ANOVA, it is easy to implement, and it provides results that are easy to interpret. 

The non-parametric statistical test that is a substitute for ANOVA is called the Kruskal-Wallis test. It shows if there is a significant difference in the means and variance of a continuous variable across three or more categorical levels when the continuous variable is not normally distributed. After the Kruskal-Wallis test, a post-hoc test called the Dunn Test, identifies the specific bivariate comparisons that significantly differ from each other.

Here is an example:
```{r}
#4a. Does daily caloric intake differ by BMI categories? What is the null hypothesis? What is the alternative hypothesis?
#the amount of calories that people intake will be the same across the BMI categories. 
#the amount of calories that people intake will have a signficant difference across the BMI categories. 
#1. Generate summary statistics of daily caloric intake by BMI cateogories?
NHANESdatabase%>%
  drop_na(DR1TKCAL)%>%
  group_by(BMIcat)%>%
  summarize(mean(DR1TKCAL), sd(DR1TKCAL), conunt=n())
#2. Generate a boxplot of daily caloric intake by BMI categories to visualize the summary statistics
ggplot(NHANESdatabase, aes(x=BMIcat, y=DR1TKCAL)) + geom_boxplot(fill="pink") + labs(x="BMI category", y="calroies consumed per day", title = "Daily calories by BMIcat")
#3. Run the Kruskal-Wallis test and the post-hoc Dunn test to see if there are any significant differences in daily caloric intake by BMI categories. The code for the Kruskal-Wallis and Dunn test is: 
#dunn.test(x=data$contvar, g=data$catvar, kw=TRUE)
dunn.test(x=NHANESdatabase$DR1TKCAL, g=NHANESdatabase$BMIcat, kw=TRUE)

```
Does the above test support or refute the alternative hypothesis? If it supports the alternative hypothesis, which categorical levels significantly differ from each other? Explain your findings.

in the case of the underweight category compared to the other categories the difference is significant, but looking at the other categories compared to one another the data shows no significant difference. 
```{r}
#4b. Does the ratio of monthly income to poverty differ by the head of household's education level? What is the null hypothesis? What is the alternative hypothesis?
#1. Generate descriptive statistics of RFITP by HHEducation levels?
NHANESdatabase%>%
  drop_na(RFITP)%>%
  group_by(HHEducation)%>%
  summarize(mean(RFITP), sd(RFITP), conunt=n())

#2. Run the Kruskal-Wallis and Dunn tests:
#dunn.test(x=data$contvar, g=data$catvar, kw=TRUE)
dunn.test(x=NHANESdatabase$RFITP, g=NHANESdatabase$HHEducation, kw=TRUE)
```
Explain your results here.

#5. Run your own Kruskal-Wallis and Dunn test below based on comparing a continuous variable across a nominal or ordinal variable with three or more levels. Show your work in the code chunk and interpret your findings.
```{r}

NHANESdatabase%>%
  drop_na(Age)%>%
  group_by(FeelBad)%>%
  summarize(mean(Age), sd(Age), conunt=n())
#2. Run the Kruskal-Wallis and Dunn tests:
#dunn.test(x=data$contvar, g=data$catvar, kw=TRUE)
dunn.test(x=NHANESdatabase$Age, g=NHANESdatabase$FeelBad, kw=TRUE)
```
in the case of the underweight category compared to the other categories the difference is significant, but looking at the other categories compared to one another the data shows no significant difference. 
(bigger than 0.05 is significant)


there is no significant difference when comparing FeelBad with Age, except with the younger categories and 0 for feel bad, which is "not at all."

